Why does female meiosis allow the opportunity for exploitation by selfish sperm? BETTER TITLE?

### 2013-09-23 writeup_and_notes/male_dependent_drive.tex
Yaniv comments on discussion on sat sept 22

\gc{Add short paragraph about sex chromosomes here? XY systems don't care. ZW systems (males ZZ, females ZW). If there is
sex ratio distortion how could this play out?}

Given the difficulty that self promiting meiotic drivers have entering the population, the speed at
which they fix if they do,
and the narrow parameter range permitting balanced polymorphisms at such loci, 
it seems very unlikely that such alleles
could drive the evolution of female suppressors of sperm-enabled
female meiotic drive.

@@

While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppression - 
	that is sperm-acting alleles that act to restore 
	fairness of female meiosis can. 
If these sperm-acting drive suppressors arise on
	the $A$ background (creating a third allele $A^{-}$) ;
	or are unlinked to the drive locus they readily invade a population segregating
	for the drive system ($AB$, see Figure). 
This allele lowers the frequency of the original drive system (perhaps to zero),
	and spreads to fixation if it does not carry strong fitness costs
	(Figure \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}A). \\
\yb{This result is unsurprising, as drive suppressors unlinked to drivers or in repulsion phase with them usually invade polymorphic drive systems (CITE).}

More surprisingly, sperm disruptors of female meiotic drive can spread
	even when tightly linked to the original drive system (B), forming
	an novel third allele $B^{-}$ as above 
	(Figures \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}B). 
$B^{-}$ displaces the original driver ($B$) by driving when not costly, 
	but by disrupting drive when transmitted to sperm it avoids low-fitness drive homozygotes ($BB^-$ and $B^-B^-$). 
Despite displacing the ancestral driver ($B$), self avoiding driver, $B^*$ it's equilibrium frequency is not always higher than that of $B$.

@@

Our results provide a plausible answer to why female oogenesis and meiosis has
not evolved to prevent sperm alleles from hijacking . 
Such self-promoting sperm alleles are usually prevented from 
entering the population if they have fitness costs, and are not
maintained as polymorphisms if they do. As such, these type of
exploitation can not drive sustained genetic conflict and so it is unlikely
that female oogenesis and meiosis will evolve to prevent their effect.

Gametogenesis and meiosis are common times for genetic conflict. This
is a consequence of the fact that as the set of alleles that consitute
an individual are in the process of parting ways, 
with only a subset of them being assured transmission.
In constrast, the perspective of the fertilizing sperm's genome is different, it has to live with the consequences of any distortion it causes 
to female meiosis through the entire next life cycle of the organism. 
It is therefore in the interest of the fertilizing sperm's genome to
make a high fitness zygote even if that means alleles in sperm
disrupting other copies of themselves from gaining a 
transmission advantage through female meiosis.

The interests of fertilizing sperm and female are quite well aligned during
syngamy. As noted above there are even opportunities for coevolution
between sperm-based supressors of female meiotic drive and the female genome.
 Thus it seems plausible that aspects of oogenesis may actively 
evolve to facilitate the influence of sperm on the outcome of female meiosis.
In light of that it is interesting to consider whether the phylogenetic variation
in the stage of oogenesis when fertilization takes place reflects not
only species' different reproductive ecologies, but also different
selection pressures shaping the interaction of sperm alleles with
female meioses.

\gc{Other pointsf}\\

 Male modification of recombination rates
POssibility that this could happen in plants if pollen emit signals to ``egg''
Discussion of OM and IN.

I THINK THERE IS A MORE SUBTLE POINT . The sperm has special knowledge
that if it allows drive it will end up in the low fitness homozygote.

>>>>

%\gc{Add short paragraph about sex chromosomes here? XY systems don't care. ZW systems (males ZZ, females ZW). If there is
%sex ratio distortion how could this play out?}

Given the difficulty that self-promoting meiotic drivers have entering the population, the speed at
which they fix if they do, and the narrow parameter range permitting balanced polymorphisms at such loci,  
it seems very unlikely that such alleles could drive the evolution of female suppressors of sperm-enabled
female meiotic drive.

@@

While  sperm enhancement of a female drive cannot displace a polymorphic female driver, sperm based drive suppression - 
	that is sperm-acting alleles that act to restore 
	fairness of female meiosis can. 
If these sperm-acting drive suppressors arise on
	the $A$ background (creating a third allele $A^{-}$) ;
	or are unlinked to the drive locus they readily invade a population segregating
	for the drive system ($AB$, see Figure). 
This allele lowers the frequency of the original drive system (perhaps to zero),
	and spreads to fixation if it does not carry strong fitness costs
	(Figure \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}A). 
\yb{This result is unsurprising, as drive suppressors unlinked to drivers or in repulsion phase with them usually invade polymorphic drive systems (CITE).}

More surprisingly, sperm disruptors of female meiotic drive can spread
	even when tightly linked to the original drive system (B), forming
	an novel third allele $B^{-}$ as above 
	(Figures \ref{Eggsperm_3_allele_cartoon} and \ref{Trajectories_of_supressors}B). 
$B^{-}$ displaces the original driver ($B$) by driving when not costly, 
	but by disrupting drive when transmitted to sperm it avoids low-fitness drive homozygotes ($BB^-$ and $B^-B^-$). 
Despite displacing the ancestral driver ($B$), self avoiding driver, $B^*$ it's equilibrium frequency is not always higher than that of $B$ \yb{(point to supp?)}.

@@

%Our results provide a plausible answer to why female oogenesis and meiosis has
%not evolved to prevent sperm alleles from hijacking . 
%Such self-promoting sperm alleles are usually prevented from 
%entering the population if they have fitness costs, and are not
%maintained as polymorphisms if they do. As such, these type of
%exploitation can not drive sustained genetic conflict and so it is unlikely
%that female oogenesis and meiosis will evolve to prevent their effect.

Sexual reproduction is a high-stakes event that determins which chromosomes are transmitted to the next generation.  
As a consequence of this intense competition for inclusion in the next generation, alleles that  gain a transmission advantage during reproduction 
	can succeed evolutionarily, even if they have bad effects on the population, 
	the mother whose offspring they will inhabbit, 
	and even the organism in which they reside. 
This potential generates numerous conflicts  \cite{genesInConflict} including sexual conflicts between mates \cite{sexualConflict}, 
	as well as conflicts between alleles that are overtransmitted in meiosis and the organisms they injure while doing so.

It seems that allowing sperm to influence the outcome of meiosis would generate a confluence of these potential conflicts -- 
	sperm could actually assist an allele that distorts female meiosis.
However, this is not the case.
We find that an allele which when in sperm help distort female meiosis in its favor, %YB [too awkward / unclear?] 
	can rarely spread through a population if it bears any cost. 
Additionally, when this self-promoting driver can spread, it cannot be maintained as a protected polymorphism, and due to its positive frequency dependence 
	it spends very little time at intermediate frequency.
As such, these type of exploitation cannot drive sustained genetic conflict and so it is unlikely
	that female oogenesis and meiosis will evolve to prevent their effect.  
This likely explains why females can delay the completion of meiosis until after fertilization 
	without exploitation by collaborations by female drivers and collaborative sperm.
%Male modification of recombination rates. \yb{trying to think through this \dots{}}

Why is it that an allele that biases female meiosis in its favor can generate a genetic conflict, but an allele in sperm that assists this female driver cannot? 
So long as the transmission advantage of female meiotic drive out way the organismal fitness cost to heterozygotes the female driver can spread when rare, and it increases in 		
	frequency until the fitness cost experienced when homozygous balances the transmission advantage.
By contrast a sperm promotor of female drive is only effective when matched with a heterozgyote female -- meaning that when rare this allele rarely enhances female drive. 
Even worse, when it does so it will find itself in a low fitness drive homozygote. 
Not only are drive-promoting sperm alleles unable to create a sustained genetic conflict, 
	but alleles in sperm with the opposite effect - that is those that prevent their own drive through female meiosis do.
This is because such drive suppressing alleles reduced their chances of forming  low fitness homozygotes.
In fact, an allele in sperm which causes the alternate allele to drive can be favored (FIG), a counterintuitive result 
	consistent with the tenuous evidence that the non-driving $ln^-$ allele is over transmitted when fertilized by $ln^+$ sperm \cite{agulnick}.

\yb{DOES THIS GO ANYWHERE? }
 We speculate that other effects of sperm on female meiosis may also be tolerated by females as a way of ensuring meiotic fairness, 
 	for example in Drosophila and other taxa, sperm accessory gland proteins (ACPs) increase the female recombination rate \cite{SOMEthING}.
Presumably this is a consequence of the resolution of double strand breaks initiated by the stressful/toxic composition of ACPs. 
However, by increasing the recombination rate sperm may decrease the effectiveness of female meiotic drivers active during MI, 
	pleiotropically enhancing female fitness in a way (while decreasing it in many others, \cite{Partridge and Hurst})

%Gametogenesis and meiosis are common times for genetic conflict. 
%This is a consequence of the fact that as the set of alleles that consitute
%	an individual are in the process of parting ways, 
%	with only a subset of them being assured transmission.
%In constrast, the perspective of the fertilizing sperm's genome is different, it has to live with the consequences of any distortion it causes 
%to female meiosis through the entire next life cycle of the organism. 
%It is therefore in the interest of the fertilizing sperm's genome to
%make a high fitness zygote even if that means alleles in sperm
%disrupting other copies of themselves from gaining a 
%transmission advantage through female meiosis.

Therefore, the interests of fertilizing sperm and female are quite well aligned during syngamy -- producing a viable and potentially fit offspring. 
%As noted above there are even opportunities for coevolution between sperm-based suppressors of female meiotic drive and the female genome.
Thus it seems plausible that aspects of oogenesis may actively evolve to facilitate the influence of sperm on ensuring a fair  outcome of female meiosis which generates fit offspring.
In light of that it is interesting to consider whether the phylogenetic variation in the stage of oogenesis when fertilization takes place reflects not
	only species' different reproductive ecologies, but also different selection pressures shaping the interaction of sperm alleles with female meioses. 
\\
\\
\\
\\
\gc{Other pointsf}\\
\\
\\
\gc{Add short paragraph about sex chromosomes here? XY systems don't care. ZW systems (males ZZ, females ZW). If there is
sex ratio distortion how could this play out?}

POssibility that this could happen in plants if pollen emit signals to ``egg''
Discussion of OM and IN.
